---
title: "R and spatial data: data preparation"
author: "Radoslaw Panczak"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    highlight: pygments
    keep_md: yes
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
---

<!-- ------------------------------------------------------------ --> 
<!-- ------------------------------------------------------------ --> 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)
library(sf)
library(tmap)
```

## Inside Airbnb

Inside Airbnb data for Melbourne obtained from http://insideairbnb.com/about.html. Verision of the data (`Date compiled`) is `07 December, 2018`.

Selected variables were kept and file was resaved as `csv` without any further modifications.

```{r}
listings <- read_csv("data/raw/inside_airbnb/listings.csv") %>% 
  select(-neighbourhood_group, - minimum_nights, -number_of_reviews, -last_review, -reviews_per_month, -calculated_host_listings_count, -availability_365)

write_csv(listings, "data/clean/listings.csv")
```

## SEIFA

### Tabular data

Socio-Economic Indexes for Areas (SEIFA) obtained from [ABS](http://www.abs.gov.au/AUSSTATS/abs@.nsf/Lookup/2033.0.55.001Main+Features12016?OpenDocument)

Data preparation included cleaning names, removing empty rows and columns and creating more sensible variable names.

```{r}
SEIFA <- read_excel("data/raw/abs/2033055001 - sa2 indexes.xls", 
                            sheet = "Table 1", skip = 5, na = "-") %>% 
  clean_names() %>%
  remove_empty(c("rows", "cols")) %>% 
  dplyr::rename(SA2_MAIN16 = `x1`, SA2_NAME16 = `x2`, 
                IRSD_s = `score_3`, IRSD_d = `decile_4`,
                IRSAD_s = `score_5`, IRSAD_d = `decile_6`, 
                IER_s = `score_7`, IER_d = `decile_8`, 
                IEO_s = `score_9`, IEO_d = `decile_10`, 
                URP = `x11`) %>% 
  dplyr::mutate(IRSD_d = as.integer(IRSD_d),
                IRSAD_d = as.integer(IRSAD_d),
                IER_d = as.integer(IER_d),
                IEO_d = as.integer(IEO_d)) %>% 
  select(-SA2_NAME16)

saveRDS(SEIFA, "data/clean/SEIFA.Rds")
```


### SA2s

SA2 files from [Australian Statistical Geography Standard (ASGS)](http://www.abs.gov.au/ausstats/abs@.nsf/mf/1270.0.55.001)

Data for GCC_NAME16 Greater Melbourne selected. SA3, GCC and STE variables removed. `SA2_MAIN16` was convertedfrom factor to numeric to ease joins.

```{r}
SA2 <- st_read("./data/raw/abs/1270055001_sa2_2016_aust_shape/SA2_2016_AUST.shp") 

SA2_2016_MELB <- SA2 %>% 
  filter(GCC_NAME16 == "Greater Melbourne") %>% 
  select(-starts_with("SA3"), -starts_with("GCC"), -starts_with("STE"), -AREASQKM16, -SA2_5DIG16)

SA2_2016_MELB$SA2_MAIN16 <- as.numeric(as.character(SA2_2016_MELB$SA2_MAIN16))

st_write(SA2_2016_MELB, "./data/clean/SA2_2016_MELB.shp")

```

