---
title: "R and spatial data: practical 2"
author: "Radoslaw Panczak"
date: 2019-02-25
output: 
  html_document: 
    highlight: pygments
    keep_md: yes
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Restarting work

## Session 

Restart session either by using GUI `Session > Restart R` or by `Ctrl + Shift + F10`. You should start with empty "Environment". Tip: you can also clear your R console with `Ctrl + L`. 

## Packages 

Load packages:

```{r message=FALSE}
if (!require("tidyverse")) install.packages("tidyverse"); library("tidyverse")
if (!require("sf")) install.packages("sf"); library("sf")
if (!require("tmap")) install.packages("tmap"); library("tmap")
if (!require("sjmisc")) install.packages("sjmisc"); library("sjmisc")
```

## Data 

Load data prepared during previous session:

```{r}
SA2_SEIFA <- readRDS("data/SA2_SEIFA.Rds")
```

# Mapping SEIFA indices in Melbourne - interactive maps

## Quick map

We already use quck map `qtm` functionality of `tmap` library. We will extend its use by providing additional argument from linked SEIFA data:

```{r message=FALSE}
tmap_mode("view")
SA2_SEIFA %>% 
  qtm(fill = "IRSAD_d")
```

We get one more map of SA2s. This time it's a thematic map! We mapped `IRSAD_d` variable to assign a colour that will fill polygon of SA2 using its particular value of decile.

## Improving interactive map

Making quick maps can quickly become cumbersome when more options need to be specified to improve the map. We will switch to building maps layer by layer, similarly to `ggplot` package. Lets rewrite the code using more advanced syntax of `tmap`. We first define our base map using `tm_shape` function and then instruct `tmap` to treat it as layers of polygons with certain attributes:


```{r}
tm_shape(SA2_SEIFA) +
  tm_polygons(col = "IRSAD_d", 
              n = 10, alpha = 0.7, palette = "RdYlGn", lwd = 0)
```

Can you guess what the parameters of `tm_polygons` function are? Use `help(tm_polygons)` to learn about the one that you don't know. Try changing a colour scheme of your map. Running code `tmaptools::palette_explorer()` will help guide your choices. 

## Faceted maps

`tm_facets` 


```{r}
tm_shape(SA2_SEIFA) +
  tm_polygons(col = c("IRSAD_d", "IER_d"), 
              n = 10, alpha = 0.7, palette = "RdYlGn", lwd = 0) +
  tm_facets(sync = TRUE, ncol = 2)
```

`col` is now `c("IRSAD_d", "IER_d")`

## Proportional symbol map


```{r message=FALSE}
tm_shape(SA2_SEIFA) +
  tm_borders(lwd = 0.5) +
  tm_bubbles(col = "IRSAD_d", size = "URP", 
             n = 10, alpha = 0.7, palette = "RdYlGn", border.lwd = 0, scale = 0.5)

```


This type of map uses size of symbols to represent numeric variable on a map.


# Mapping SEIFA indices in Melbourne - static maps

Interactive maps are great for exploration. On other occasions ‘static’ map can be a better option, for instance when you want to include it in printed report. We will learn how to create one with slightly more advanced use of `tmap` library functions.

## Quick map

```{r message=FALSE}
tmap_mode("plot")
```


```{r}
tm_shape(SA2_SEIFA) +
  tm_polygons("IRSAD_d", title = "IRSAD deciles", 
              palette = "RdYlGn", n = 10, lwd = 0) 
```


First we instructed tmap package to change from interactive mode into static mode with `tmap_mode()` function. Notice that rest of parameters is the same as in interactive mode.


## Improving static map

```{r}
tm_shape(SA2_SEIFA) +
  tm_polygons("IRSAD_d", title = "IRSAD deciles", 
              palette = "RdYlGn", n = 10, lwd = 0,
              legend.hist = T) +
  tm_scale_bar(position = c("right", "bottom")) +
  tm_compass(position = c("left", "top"))+
  tm_layout(frame = F, 
            main.title = "IRSAD deciles of SA2 areas in Melbourne",
            main.title.size = 1.25, 
            main.title.position = c("left", "top"),
            legend.hist.size = 0.5, 
            legend.outside = T)
```


We have built a map layer by layer, calling different functions and specifying various arguments. We also improved the cartography by providing legend and a histogram (a graph of distribution of values). 
We specified our data source in `tm_shape()` function; many parameters indicating how polygons are displayed are specified in `tm_polygons()` function; we also added north arrow and scale with functions names appropriately; finally `tm_layout` helped us to put all small details together as legend and title). 

Last but not least we saved our map can be saved as file on the disk. That can be done with a call `tmap_save(tmap_last(), "path_to_save/name.png")`. Notice that we used function inside a finction there to use last displayed map.


# Further topics

1. Try mapping indices other than `IRSAD`.

2. Try creating static map with proportianl symbols.

3. Try creating static maps with facets.

