---
title: "R and spatial data: intermediate 1"
author: "Radoslaw Panczak"
date: 2019-02-26
output: 
  html_document: 
    # css: custom.css
    toc: true
    toc_depth: 4
    toc_float: true
    number_sections: true
    theme: united 
    highlight: pygments 
---

# Maintenance 

## Restart R session in Rstudio

Use keyboard shortcut `Ctrl + Shift + F10`

Check environment

Explain why

## Load the libraries

```{r}
set.seed(1234)
library(tidyverse) # data manipulation and graphs
library(sf) # simple features to handle geographical data
library(tmap) # mapping library
library(skimr) # summarizing data
```

# Your third geovizualization - mapping Airbnb in Melbourne

# Background information

[Inside Airbnb](http://insideairbnb.com/about.html) is website that describes itself as:

> an independent, non-commercial set of tools and data that allows you to explore how Airbnb is really being used in cities around the world. By analyzing publicly available information about a city's Airbnb's listings, Inside Airbnb provides filters and key metrics so you can see how Airbnb is being used to compete with the residential housing market.

The dataset is not offically approved by Airbnb but is released under open license and in the absence of other data - was used by many research projects around the world.

# Data input

## Read the data

```{r}
listings <- read_csv("./data/clean/listings.csv")
```

## Examine the data

```{r}
slice(listings, 1:5)
```

```{r, results='asis' }
skim(listings) %>% skimr::kable()
```

## Turning data spatial

http://spatialreference.org/ref/epsg/wgs-84/ 

```{r}
listings_sf = st_as_sf(listings, coords = c("longitude", "latitude"), crs = 4326)
```

Look at the difference

```{r}
slice(listings_sf, 1:5)
```

## Your first map

```{r}
tmap_mode("view")
qtm(listings_sf)
```

## Improving your map

Map variable in your data to visual encoding on map.

```{r}
ggplot(listings_sf, aes(price)) + geom_histogram()
```

```{r}
summary(listings_sf$price)
```

```{r}
listings_sf %>% 
  sample_n(1000) %>%
  tm_shape() + 
  tm_dots(col = "price", style = "quantile", n = 10, palette = "RdYlGn")
```

`tmaptools::palette_explorer()`


## Static map

```{r}
tmap_mode("plot")
qtm(listings_sf)
```

## Improving your map

```{r}
listings_sf %>% 
  # sample_n(1000) %>%
  tm_shape() + 
  tm_dots(col = "price", title = "Price of listing",
          style = "quantile", n = 10, 
          palette = "RdYlGn", legend.hist = T) +  
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass(position = c("left", "top"))+
  tm_layout(frame = F, 
            main.title = "Price of Airbnb listing in Melbourne",
            main.title.size = 1.25, 
            main.title.position = c("left", "top"),
            legend.hist.size = 0.5, 
            legend.outside = T)
```

```{r}
listings_sf %>% 
  ggplot(aes(room_type)) + geom_bar()
```

```{r}
listings_sf %>% 
  ggplot(aes(room_type, price)) + geom_boxplot() + scale_y_log10()
```

```{r}
listings_sf %>% 
  filter(room_type != "Shared room") %>% 
  # sample_n(1000) %>%
  tm_shape() + 
  tm_dots(col = "price", title = "Price of listing",
          style = "quantile", n = 10, 
          palette = "RdYlGn", legend.hist = T) +  
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass(position = c("left", "top"))+
  tm_layout(frame = F, 
            main.title = "Price of Airbnb listing in Melbourne",
            main.title.size = 1.25, 
            main.title.position = c("left", "top"),
            legend.hist.size = 0.5, 
            legend.outside = T) +
  tm_facets(by = "room_type")

```


# Resources

Short video introducing projections https://www.youtube.com/watch?v=kIID5FDi2JQ 

Learn more about simple features and `sf` library https://r-spatial.github.io/sf/

Data Carpentry’s Geospatial Workshop http://datacarpentry.org/geospatial-workshop/




