---
title: "R and spatial data: intermediate 2"
author: "Radoslaw Panczak"
date: 2019-02-26
output: 
  html_document: 
    highlight: pygments
    keep_md: yes
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
---

# Maintenance 

## Restart R session in Rstudio

Use keyboard shortcut `Ctrl + Shift + F10`

## Load the libraries 

```{r message=FALSE}
set.seed(1234)

if (!require("tidyverse")) install.packages("tidyverse"); library("tidyverse")
if (!require("sf")) install.packages("sf"); library("sf")
if (!require("tmap")) install.packages("tmap"); library("tmap")
if (!require("spdep")) install.packages("spdep"); library("spdep")
if (!require("sjmisc")) install.packages("sjmisc"); library("sjmisc")

tmap_mode("view")
```

## Load the data

```{r}
# prepared earlier
SA2_SEIFA <- readRDS("data/SA2_SEIFA.Rds")
```

# From `sf` to `sp` 

So far we worked with spatial objects represented by simple features from `sf` library. Long before `sf` library appeared R used different representation of spatial objects provided by `sp` library. Many packages might still depend on it so we will learn how to transform the data between then.

One notable example is `spdep` package that offers very powerful se tof tools to analyse spatial patterns of data.

We will reproject our SA2 & SEIFA data prepared before and convert it to `sp` object using `as` function:

```{r}
SA2_SEIFA_proj <- SA2_SEIFA %>% 
  st_transform(3112) %>% 
  filter(!is.na(IRSAD_s)) %>% 
  as("Spatial")
```

Our object is now of a different class:

```{r}
class(SA2_SEIFA_proj)
```

Because of that it has different properties and will behave differently. For instance. We cannot simply view the data associated with polygons:

```{r}
View(SA2_SEIFA_proj)
```

Instead of that we have to access so called data "slot" by using `@` notation:
```{r}
View(SA2_SEIFA_proj@data)
```

Luckily, `tmap` library can work with both `sf` and `sp` objects so all we have learned so far can be still used. Try creating a quick map (`qtm` function) with `SA2_SEIFA_proj`.

# Spatial neighbours

In order to examine relationship between regions we have to define how we would like to define who is included as a neighbour. One common approach is to define "queen" type of neighbouring (queen being relation to the queen's movement in chess). We can create neighbours list from polygon list with `poly2nb` function:

```{r}
(SA2_weights_queen <- poly2nb(SA2_SEIFA_proj, queen = TRUE))
```

We can now turn this object into spatial weights (`listw` class) that are required by some tools:

```{r}
(SA2_weights_queen_list <- nb2listw(SA2_weights_queen, style="W", zero.policy=TRUE))
```

Picture might give it better overview than description - let's plot the object to see how polygons are connected:

```{r}
plot(SA2_SEIFA_proj, border = "grey60")
plot(SA2_weights_queen, coordinates(SA2_SEIFA_proj), pch = 19, cex = 0.6, add = TRUE)
```

We now defined all polygons that touch any part of their boundary to become neighbours. We will now work with statistics that use this neighbours to determine degree of global and local clustering.

# Global spatial clustering

In order to check if our data is clustered in space we could use Moran's I test for spatial autocorrelation. `moran.test` function from `spdep` package will perform that:

```{r}
(SA2_IRSAD_s_Moran <- moran.test(SA2_SEIFA_proj$IRSAD_s, listw = SA2_weights_queen_list))
```

Interrpretation of the statistic is as follow:

>Values of I usually range from −1 to +1. Values significantly below -1/(N-1) indicate negative spatial autocorrelation and values significantly above -1/(N-1) indicate positive spatial autocorrelation. 

The significance of the statistic can be obtained from the Monte Carlo permutation test:

```{r}
(SA2_IRSAD_s_Moran_MC <- moran.mc(SA2_SEIFA_proj$IRSAD_s,
                                 SA2_weights_queen_list,
                                 nsim = 999))
```

We can also get some insights into spatial patterning of the data by using Moran scatterplot, which plots data against its spatially lagged values, 

```{r}
moran.plot(SA2_SEIFA_proj$IRSAD_s, listw = nb2listw(SA2_weights_queen),
           xlab = "ISRAD score", ylab = "Lag of ISRAD score")
```

Data points are divided into four "quadrants" depending on where they lie in respect to their values to lagged values relation. For instance top-right quadrant contains higher values of variable surrounded also by neighboring areas with higher values as well. That could point to spatial clustering. Bottom-right quadrant on the other hand points to high values of variable surrounded by areas with lower values - indication for spatial outliers. We will use these quadrants and visualize them on maps. 

# Local spatial clustering

Moran's I tells us if our data are globally clustered without indicating where exactly it is happening. We can learn about location of clusters and outliers by using local indicators of spatial association (LISA).

## Data preparation 

We first prepare our data by creating a standardized version of it using `scale` function. 

```{r}
SA2_SEIFA_proj$IRSAD_s_std <- scale(SA2_SEIFA_proj$IRSAD_s) 
```

We also prepare spatially lagged variable using spatial weights we created above.

```{r}
SA2_SEIFA_proj$IRSAD_s_lag <- lag.listw(SA2_weights_queen_list, SA2_SEIFA_proj$IRSAD_s_std)
```

We can create a variable with four classes depending on which quadrant of the Moran's plot the region lies on the scaterplot of standardized versus lagged values. Four possible conditions are:

```{r}
SA2_SEIFA_proj$quad <- NA
SA2_SEIFA_proj@data[SA2_SEIFA_proj$IRSAD_s_std > 0 & SA2_SEIFA_proj$IRSAD_s_lag > 0, "quad"] <- 1
SA2_SEIFA_proj@data[SA2_SEIFA_proj$IRSAD_s_std < 0 & SA2_SEIFA_proj$IRSAD_s_lag < 0, "quad"] <- 2
SA2_SEIFA_proj@data[SA2_SEIFA_proj$IRSAD_s_std > 0 & SA2_SEIFA_proj$IRSAD_s_lag < 0, "quad"] <- 3
SA2_SEIFA_proj@data[SA2_SEIFA_proj$IRSAD_s_std < 0 & SA2_SEIFA_proj$IRSAD_s_lag > 0, "quad"] <- 4
```

Now we are ready to map our results. Each region belongs to one quadrant:

```{r}
frq(SA2_SEIFA_proj$quad)
```

We can map it into four categories giving it descriptive names and colours matching clusters (reds) and outliers (blues):

```{r}
tm_shape(SA2_SEIFA_proj) +
  tm_fill(col = "quad", alpha = 0.5, style = "cat",
          palette = c("red", "blue", "lightpink", "skyblue2"),
          labels = c("High-High", "Low-Low", "High-Low", "Low-High"),
          title = "Local Moran's I")
```

## LISA

LISA analysis can be performed using `localmoran` function from `spdep` package. We use variable of interest (`IRSAD_s` score) and `listw` object with spatial weights of neighbours:

```{r}
SA2_IRSAD_s_lm <- localmoran(SA2_SEIFA_proj$IRSAD_s, listw = SA2_weights_queen_list)
```

We obtain an object of localmoran class that stores five results:

```{r}
summary(SA2_IRSAD_s_lm)
```

The values here indicate:

*Ii* - local moran statistic
*E.Ii* - expectation of local moran statistic
*Var.Ii* - variance of local moran statistic
*Z.Ii* - standard deviate of local moran statistic
*Pr()* - p-value of local moran statistic

We can now extend quadrant assignment by additionally marking results significant depending on p-value obtained from the `localmoran` test (stored in the fifth column of results).

```{r}
SA2_SEIFA_proj$quad_sig <- SA2_SEIFA_proj$quad
SA2_SEIFA_proj@data[SA2_IRSAD_s_lm[, 5] >= 0.05, "quad_sig"] <- 5  
```

Taking itnto account significance of results, we can add one more category to the results for regions that are not significantly different according to permutation test:

```{r}
frq(SA2_SEIFA_proj$quad_sig)
```

Since we only have first and second category, we modify our `palette` and `labels` arguments and also remove colour for insignificat results:

```{r}
tm_shape(SA2_SEIFA_proj) +
  tm_fill(col = "quad_sig", alpha = 0.5, style = "cat",
          # palette = c("red", "blue", "lightpink", "skyblue2", "white"), 
          # labels = c("High-High", "Low-Low", "High-Low", "Low-High", "Not Signif."), 
          palette = c("red", "blue", NA), 
          labels = c("High-High", "Low-Low", "Not Signif."), 
          title = "Local Moran's I")
```

How does that map compare to original map of IRSAD score and deciles?

# Further topics

1. Construct neighbours using argument `queen = FALSE`. What differences can you see. In what types of situations would you expect it boe the biggest issue?

2. Examine global and local autocorrelation of other SEIFA index. 

