---
title: "R and spatial data: practical 2"
author: "Radoslaw Panczak"
date: 2019-02-25
output: 
  html_document: 
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Restarting work

## Session 

Restart session either by using GUI `Session > Restart R` or by `Ctrl + Shift + F10`. You should start with empty "Environment". Tip: you can also clear your R console with `Ctrl + L`. 

## Packages 

Load packages:

```{r message=FALSE}
if (!require("tidyverse")) install.packages("tidyverse"); library("tidyverse")
if (!require("sf")) install.packages("sf"); library("sf")
if (!require("tmap")) install.packages("tmap"); library("tmap")
if (!require("tmaptools")) install.packages("tmaptools"); library("tmaptools")
if (!require("sjmisc")) install.packages("sjmisc"); library("sjmisc")
```

## Data 

Load data prepared during previous session:

```{r}
SA2_SEIFA <- readRDS("data/SA2_SEIFA.Rds")
```

# Mapping SEIFA indices in Melbourne - interactive maps

## Quick map

We already know how to use quck map `qtm` functionality of `tmap` library. We will extend its use by providing additional argument from linked SEIFA data:

```{r message=FALSE}
tmap_mode("view")
SA2_SEIFA %>% 
  qtm(fill = "IRSAD_d")
```

We get one more map of SA2s. This time it's a thematic map! We mapped `IRSAD_d` variable to assign a colour that will fill polygon of SA2 using its particular value of decile.

## Improving interactive map

Making quick maps can quickly become cumbersome when more options need to be specified to improve the map. We will switch to building maps layer by layer, similarly to `ggplot` package. Lets rewrite the code using more advanced syntax of `tmap`. We first define our base map using `tm_shape` function and then instruct `tmap` to treat it as layers of polygons with certain attributes:


```{r}
tm_shape(SA2_SEIFA) +
  tm_polygons(col = "IRSAD_d", 
              n = 10, alpha = 0.7, palette = "RdYlGn", lwd = 0)
```

Can you guess what the arguments of `tm_polygons` function are? Use `help(tm_polygons)` to learn about the one that you don't know. Try changing a colour scheme of your map. Running code `tmaptools::palette_explorer()` will help guide your choices. 

## Faceted maps

Sometimes it might be useful to explore two variables at once. `tmap` lets you create facets (in a similar way as `ggplot` does) using `tm_facets`. We can create linked interactive display of two variables:

```{r}
tm_shape(SA2_SEIFA) +
  tm_polygons(col = c("IRSAD_d", "IER_d"), 
              n = 10, alpha = 0.7, palette = "RdYlGn", lwd = 0) +
  tm_facets(sync = TRUE, ncol = 2)
```

Pay attention that the `col` argumant consists now of two variables combines using function call `c("IRSAD_d", "IER_d")`.

## Proportional symbol map

Variables can be mapped not only to colours, but also to size. We replace `tm_polygons` call with `tm_bubbles` to get a proportional symbol map:

```{r message=FALSE}
tm_shape(SA2_SEIFA) +
  tm_borders(lwd = 0.5) +
  tm_bubbles(col = "IRSAD_d", size = "URP", 
             n = 10, alpha = 0.7, palette = "RdYlGn", border.lwd = 0, scale = 0.5)

```

Note that the `size` argument has been mapped to `URP` variable that contains values of usual resident population in the area. In this way - the size of bubbles is now determinded by population. Can you think of advantages of such map? Are there any locations where it can be useful?

# Mapping SEIFA indices in Melbourne - static maps

Interactive maps are great for exploration. On other occasions ‘static’ map can be a better option, for instance when you want to include it in a printed report. We will learn how to create one with slightly more advanced use of `tmap` library functions.

## Quick map

First we change mode of `tmap`:

```{r message=FALSE}
tmap_mode("plot")
```

Now we can once again create a choropleth map:

```{r}
tm_shape(SA2_SEIFA) +
  tm_polygons("IRSAD_d", title = "IRSAD deciles", 
              palette = "RdYlGn", n = 10, lwd = 0) 
```

Notice that the arguments are the same as in interactive mode.

## Improving static map

Our map is already starting to look good. We can improve some aspects of design by adding and arranging items of the layout such as scale bar, north arrow and creating histogram of data according to classes used for mapping: 

```{r}
tm_shape(SA2_SEIFA) +
  tm_polygons("IRSAD_d", title = "IRSAD deciles", 
              palette = "RdYlGn", n = 10, lwd = 0,
              legend.hist = T) +
  tm_scale_bar(position = c("right", "bottom")) +
  tm_compass(position = c("left", "top"))+
  tm_layout(frame = F, 
            main.title = "IRSAD deciles of SA2 areas in Melbourne",
            main.title.size = 1.25, 
            main.title.position = c("left", "top"),
            legend.hist.size = 0.5, 
            legend.outside = T)
```

We have built a map layer by layer, calling different functions and specifying various arguments. We also improved the cartography by providing legend and a histogram (a graph of distribution of values). 
We specified our data source in `tm_shape()` function; many arguments indicating how polygons are displayed are specified in `tm_polygons()` function; we also added north arrow and scale with functions names appropriately; finally `tm_layout` helped us to put all small details together as legend and title). 

Last but not least we saved our map can be saved as file on the disk. That can be done with a call `tmap_save(tmap_last(), "path_to_save/name.png")`. Notice that we used function inside a finction there to use last displayed map.


# Further topics

1. Try mapping indices other than `IRSAD`.

2. Try creating static map with proportianl symbols.

3. Try creating static maps with facets.

